Reflections on reflection for \\ intuitionistic set theories

### 2019-01-10 paper-reflection.tex
Optimality result

The basic form of the reflection principle for Zermelo--Fraenkel set
theory~$\ZF$ is the following.
\begin{thm}Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\ZF$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$, obtained by
substituting any occurence of~``$\forall x$'' and~``$\exists x$'' by~``$\forall
x \in S$'' and~``$\exists x \in S$''.
Furthermore, the resulting set~$S$ may be supposed to be transitive, closed
under subsets or even to be a stage~$V_\alpha$ of the cumulative hierarchy.
% XXX: "hence" instead of "or even"?
\end{thm}

@@

\begin{thm}\label{thm:zf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\ZF$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
\end{thm}

@@

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
\end{thm}

@@

@@

>>>>

The basic form of the reflection principle for Zermelo--Fraenkel set
theory~$\ZF$ is the following.
\begin{thm}Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\ZF$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$, obtained by
substituting any occurence of~``$\forall x$'' and~``$\exists x$'' by~``$\forall
x \in S$'' and~``$\exists x \in S$''.
Furthermore, the resulting set~$S$ may be supposed to be transitive, to be closed
under subsets or even to be a stage~$V_\alpha$ of the cumulative hierarchy;
% XXX: "hence" instead of "or even"?
and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

@@

\begin{thm}\label{thm:zf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\ZF$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$. Furthermore, the
resulting set~$S$ may be supposed to be transitive and to be closed under
subsets; and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

@@

\begin{thm}\label{thm:izf-bigstep}
Let~$\varphi(x_1,\ldots,x_n)$ be a formula in the language of set
theory with free variables as indicated. Then~$\IZF+\RRS$ proves
\[ \forall M\_
  \exists S \supseteq M\_
  \forall x_1,\ldots,x_n \in S\_
  \varphi(x_1,\ldots,x_n) \Leftrightarrow \varphi^S(x_1,\ldots,x_n), \]
where~$\varphi^S$ is the~$S$-relativization of~$\varphi$.
Furthermore, the resulting set~$S$ may be supposed to be transitive and to be closed
under subsets; and given not a single formula~$\varphi$ but a finite
list~$\varphi_1,\ldots,\varphi_s$ of formulas, we may suppose that~$S$ reflects
all of them.
\end{thm}

@@

\begin{thm}\label{thm:refl-entails-rrs}
Over~$\IZF$, each instance of Aczel's relation reflection scheme~$\RRS$ can be
deduced from suitable instances of the assumption that, given a finite list
of formulas, for every set~$M$ there is a set~$S \supseteq M$ reflecting the
given formulas.
\end{thm}

\begin{proof}We verify the equivalent~$\MDC$ instead of~$\RRS$. Let~$X$ and~$R
\subseteq X \times X$ be classes. Let~$A \subseteq X$ be a subset and
suppose~$\forall x \in X\_ \exists y \in X\_ \langle x,y \rangle \in R$.

By assumption, there is a set~$S \supseteq A$ which reflects the three
formulas~``$x \in X$'', ``$\langle x,y \rangle \in R$'' and~``$\exists y \in X\_ \langle
x,y \rangle \in R$''.

By separation, the class~$S \cap X$ is a set, and hence we may set~$f(k) \defeq S \cap X$
for all numbers~$k$. Then~$A \subseteq f(0)$, any given~$x \in f(k)$, by
assumption and the reflecting property of~$S$ there exists~$y \in S$ such
that~$y \in X$ (hence~$y \in f(k+1)$) and~$\langle x,y \rangle \in R$.
\end{proof}

@@

% XXX: Think about pairs.
